---
title: "An automated procedure for phenology data analyses"
author: "Lea Rosa Maria Lipphardt"
date: "2024-11-19"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(repos = c(CRAN = "https://cran.rstudio.com/"))

```

## Introduction
 
This automated procedure streamlines analysis of phenology data,
enabling more efficient estimation of key metrics such as hatching
dates, breeding success, and chick survival. The data is sourced from
the ACCESS database, NEMO, and processed through a series of R scripts.

The workflow is structured as follows:

1.  Setup
    1.1 Confirm system architecture (32-bit or 64-bit)
    1.2 Install and load the required packages
2.  Import Data from Access
    2.1 Connect to the Access database using ODBC
    2.2 Retrieve data via SQL queries
    2.3 Control section
3.  Data Transformation and Preperation
    3.1 Format dates and optionally subset data
    3.2 For Svarthamaren assign a specific breeding period
4.  Flag and Document Exceptions
    4.1 Identify and document data anomalies
5.  Phenology Estimates    
6.  Estimate Hatching Dates
    5.1 Calculate hatching dates and assess accuracy
7.  Chick Presence and Breeding Success
8.  Summarize Findings

This document is written in R Markdown, a versatile format for creating
HTML, PDF, and Word documents. For more details on R Markdown, visit
<http://rmarkdown.rstudio.com>.

To execute the R code, click the green arrow in the Script Editor
(Source pane). Press the Knit button to generate a final document that
integrates both narrative content and the output from your R code
chunks.

## 1. Setup

To ensure compatibility when connecting to databases such as Microsoft
Access, ensure the architecture (32-bit or 64-bit) of the following
components match:

-   **R Version**

    Check your R bit version by running:

```{r}
R.version$arch
```

    The output will indicate either x86_64 (64-bit) or i386 (32-bit).

-   **ODBC Driver**

    -   nstall the ODBC driver matching your R version:

        -   For 32-bit R: Install the 32-bit ODBC driver.

        -   For 64-bit R: Install the 64-bit ODBC driver.

    -   Download drivers from the [Microsoft ODBC Driver
        website](https://learn.microsoft.com/en-us/sql/connect/odbc/download-odbc-driver-for-sql-server).

-   **Microsoft Office**

    -   Check your Microsoft Office version:

        -   Open any Office app (e.g., Word), go to **File** \>
            **Account** \> **About**, and look for "32-bit" or "64-bit."

    -   If your Office version doesn't match your R version, consider
        installing the Access Database Engine for the appropriate bit
        version: [Download Access Database
        Engine](https://www.microsoft.com/en-us/download/details.aspx?id=54920).




Install and load the required packages

```{r, results='hide', message=FALSE, warning=FALSE}

# Package Management automatically installs and loads the packages if they are not already available
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(odbc, DBI, tidyverse, lubridate, dplyr)

```

## 2. Import Data
### 2.1 Connect to the Access database using ODBC

```{r, results='hide', message=FALSE, warning=FALSE}
# Define the function to connect to Access
connect_to_access <- function(db_path) {
  dbConnect(odbc::odbc(), 
            .connection_string = paste0("Driver={Microsoft Access Driver (*.mdb, *.accdb)};DBQ=", db_path))
}

# Define the database path
db_path <- "C:/Users/lea.lipphardt/Documents/NEMOx/Spitsbergen_2024.mdb"

# Connect to the Access database
con <- connect_to_access(db_path)

```

### 2.2 Define the SQL query

2.1 Create dataframe using Structured query language (SQL)

**SELECT** lets you choose the specific columns (from the different
tables) that you want to use.

**AS** allows you to assign a temporary name (alias) to tables and
columns. In the R script, aliases are used for the tables to make the
SQL query more readable.

**ON** specifies the condition for the join (eg. v.NestStatus =
nc.NestContentID) which tables store the same information under which
columns.

**INNER JOIN** lets you merge tables, ensuring that only records that
have matching values in both tables are included.

You can alter the code if you for instance want to include more columns
with adding the Table.Column to SELECT or further tables using INNE JOIN
(Table.Column ON Table.Column).

```{r}
# Define the SQL query
query <- "
SELECT s.spcENG, n.Locality, n.Area, n.NestNumber, v.VisitDate, nc.NestContentText, v.NestClutch
FROM (((Visits AS v
INNER JOIN NestContent AS nc ON v.NestStatus = nc.NestContentID)
INNER JOIN Nest AS n ON n.NestUniqueID = v.NestUniqueID)
INNER JOIN Species AS s ON n.Species = s.EUnr)
"

# Execute the query and fetch results into a dataframe
df_selected <- dbGetQuery(con, query)

# Inspect the fetched data
head(df_selected)
str(df_selected)

# Disconnect from the database
dbDisconnect(con)

```

### 2.3 Control section
No need to run this section
```{r}
nrow(df_selected)

# Count the number of occurrences of each unique character in NestClutch
table(df_selected$NestClutch)

# Convert df_selected into a tibble and summarize with dplyr
df_selected %>%
  as_tibble() %>%
  count(NestClutch)
```
## 3. Data Transformation and Preperation
### 3.1 Convert dates into the right format and optional subset of data

```{r}
# Convert VisitDate column to Date format (ensures consistent handling of date values)
df_selected$VisitDate <- as.Date(df_selected$VisitDate)

#Extract the breeding season year:
  # For Svalbard: BreedingSeason is represented as YYYY (e.g., 2024)
  # For Svarthamaren: BreedingSeason will later be adjusted to YYYY/YYYY 
df_selected$BreedingSeason <- as.numeric(format(df_selected$VisitDate, "%Y"))

# Subset the data for a specific BreedingSeason, specie, location
df_selected_BreedingSeason <- subset(df_selected, BreedingSeason == "2024")
#df_selected_specie <- subset(df_selected_BreedingSeason, spcENG == "Black-legged kittiwake")
#df_selected_location <- subset(df_selected_specie, Locality == "Ossian Sarsfjellet")

```

### 3.2 For Svarthamaren run the bellow to assign a specific breeding period

Define the breeding season start and end dates relative to each BreedingSeason for Svarthamaren.
The BreedingSeason will be YYYY for NestNumbers from Svalbard and YYYY/YYYY for Svarthamaren

```{r}
# For Svarthamaren run the bellow to assign a specific breeding period
# Define the breeding season start and end dates relative to each BreedingSeason
df_selected$BreedingSeason <- ifelse(
  format(df_selected$VisitDate, "%m-%d") >= "11-01", 
  # If the VisitDate is in November or December, use current BreedingSeason/current BreedingSeason + 1
  paste0(df_selected$BreedingSeason, "/", df_selected$BreedingSeason + 1),
  ifelse(
    format(df_selected$VisitDate, "%m-%d") <= "02-29", 
    # If the VisitDate is in January or February, use previous BreedingSeason/current BreedingSeason
    paste0(df_selected$BreedingSeason - 1, "/", df_selected$BreedingSeason),
    NA # Otherwise, not in the breeding season
  )
)
```

## 4 Flag and Document Exceptions
### 4.1 Only E/C/empty NestClutch values
NestClutch values that only include egg/chick/empty entries for a NestNumber for a BreedingSeason will get a Note appended as defined bellow
```{r}

append_note_to_clutch_values <- function(df) {
  # Define regex patterns for different categories
  clutch_values_e <- "^([1-9][Ee]|[1-9][Ee],[1-9][Ee])$"
  clutch_values_c <- "^([1-9][Cc]|[1-9][Cc],[1-9][Cc])$"
  
  # Define valid clutch values for empty entries
  empty_clutch_values <- c("")
  
  # Initialize the Note column with empty strings
  df$Note <- ""
  
  # Iterate over unique NestNumbers and BreedingSeasons
  for (nest in unique(df$NestNumber)) {
    for (BreedingSeason in unique(df$BreedingSeason)) {
      # Subset the data for the current NestNumber and BreedingSeason
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, ]
      
      # Remove rows with NA values in NestClutch
      subset_df <- subset_df[!is.na(subset_df$NestClutch), ]
      
      # Check conditions using regex and append notes
      if (all(grepl(clutch_values_e, subset_df$NestClutch))) {
        df$Note[df$NestNumber == nest & df$BreedingSeason == BreedingSeason] <- "Only E-entries"
      } else if (all(grepl(clutch_values_c, subset_df$NestClutch))) {
        df$Note[df$NestNumber == nest & df$BreedingSeason == BreedingSeason] <- "Only C-entries"
      } else if (all(subset_df$NestClutch %in% empty_clutch_values)) {
        df$Note[df$NestNumber == nest & df$BreedingSeason == BreedingSeason] <- "Only empty-entries"
      }
    }
  }
  
  return(df)
}
```

### 4.2 False NestClutch entries
Apply the function to the data for appending a Note where you expect a false entry 
for the NestClutch value
```{r}
df_selected_BreedingSeason <- append_note_to_clutch_values(df_selected_BreedingSeason)

# Append note to exceptions
# This flags exceptions in the data based on specific conditions:
#   "False_E": Occurs when a clutch value of 'E' is recorded after a 'C' entry.
#   "False_C": Occurs when a clutch value of 'C' is recorded and followed by two consecutive 'E' entries.

# Define regex patterns for clutch values
clutch_value_e <- "^([1-9][Ee]|[1-9][Ee],[1-9][Ee])$"  # Possible combinations for egg entries
clutch_value_c <- "^([1-9][Cc]|[1-9][Cc],[1-9][Cc])$"  # Possible combinations for chick entries

# Apply the logic to flag False_E for '1E' entries following a '1C' entry
df_selected_BreedingSeason <- df_selected_BreedingSeason %>%
  group_by(BreedingSeason, NestNumber) %>%
  arrange(VisitDate, .by_group = TRUE) %>%
  mutate(
    Note = case_when(
      grepl(clutch_value_e, NestClutch) & lag(grepl(clutch_value_c, NestClutch), default = FALSE) ~ "False_E",
      TRUE ~ Note # Retain existing notes for other cases
    )
  )


# Apply the logic to flag False_C for '1C' entries followed by two '1E' entries
df_selected_BreedingSeason <- df_selected_BreedingSeason %>%
  group_by(BreedingSeason, NestNumber) %>%
  arrange(VisitDate, .by_group = TRUE) %>%
  mutate(
    Note = case_when(
      grepl(clutch_value_c, NestClutch) & 
        lead(grepl(clutch_value_e, NestClutch), n = 1, default = FALSE) &
        lead(grepl(clutch_value_e, NestClutch), n = 2, default = FALSE) ~ "False_C",
      TRUE ~ Note # Retain existing notes for other cases
    )
  )
```

## 5. Phenology Estimates
```{r}

# Find first visit with egg
find_first_E <- function(df) {
  clutch_values_e <- "^([1-9][Ee]|[1-9][Ee],[1-9][Ee])$"
  
  df$First_E <- as.Date(NA)
  
  for (nest in unique(df$NestNumber)) {
    for (BreedingSeason in unique(df$BreedingSeason)) {
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, ]
      subset_df$VisitDate <- as.Date(subset_df$VisitDate)
      
      matching_clutches <- subset_df[grepl(clutch_values_e, subset_df$NestClutch), ]
      
      if (nrow(matching_clutches) > 0) {
        first_entry <- matching_clutches[which.min(matching_clutches$VisitDate), ]
        df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, "First_E"] <- first_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Find last visit with egg
find_last_E <- function(df) {
  clutch_values_e <- "^([1-9][Ee]|[1-9][Ee],[1-9][Ee])$"
  
  df$Last_E <- as.Date(NA)
  
  for (nest in unique(df$NestNumber)) {
    for (BreedingSeason in unique(df$BreedingSeason)) {
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, ]
      subset_df$VisitDate <- as.Date(subset_df$VisitDate)
      
      matching_clutches <- subset_df[grepl(clutch_values_e, subset_df$NestClutch), ]
      
      if (nrow(matching_clutches) > 0) {
        last_entry <- matching_clutches[which.max(matching_clutches$VisitDate), ]
        df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, "Last_E"] <- last_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Find first visit with chick
find_first_C <- function(df) {
  clutch_values_combined <- "^([1-9][EeCc]|[1-9][EeCc],[1-9][EeCc])$"
  
  df$First_C <- as.Date(NA)
  
  for (nest in unique(df$NestNumber)) {
    for (BreedingSeason in unique(df$BreedingSeason)) {
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, ]
      subset_df$VisitDate <- as.Date(subset_df$VisitDate)
      
      matching_clutches <- subset_df[grepl(clutch_values_combined, subset_df$NestClutch), ]
      
      if (nrow(matching_clutches) > 0) {
        first_entry <- matching_clutches[which.min(matching_clutches$VisitDate), ]
        df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, "First_C"] <- first_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Find last visit with chick
find_last_C <- function(df) {
  clutch_values_combined <- "^([1-9][EeCc]|[1-9][EeCc],[1-9][EeCc])$"
  
  df$Last_C <- as.Date(NA)
  
  for (nest in unique(df$NestNumber)) {
    for (BreedingSeason in unique(df$BreedingSeason)) {
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, ]
      subset_df$VisitDate <- as.Date(subset_df$VisitDate)
      
      matching_clutches <- subset_df[grepl(clutch_values_combined, subset_df$NestClutch), ]
      
      if (nrow(matching_clutches) > 0) {
        last_entry <- matching_clutches[which.max(matching_clutches$VisitDate), ]
        df[df$NestNumber == nest & df$BreedingSeason == BreedingSeason, "Last_C"] <- last_entry$VisitDate
      }
    }
  }
  
  return(df)
}

find_last_visit <- function(df) {
  # Create new columns initialized with NA
  df$LastOfVisitDate <- as.Date(NA)
  df$LastStatus_All <- NA
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (breedingseason in unique(df$BreedingSeason)) {
      # Subset the data for the current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$BreedingSeason == breedingseason, ]
      
      # Find the last visit based on VisitDate
      if (nrow(subset_df) > 0) {
        last_entry <- subset_df[which.max(subset_df$VisitDate), ]
        
        # Assign the Last_Visit and LastStatus_All values to all rows for this NestNumber and Year
        df[df$NestNumber == nest & df$BreedingSeason == breedingseason, "LastOfVisitDate"] <- last_entry$VisitDate
        df[df$NestNumber == nest & df$BreedingSeason == breedingseason, "LastStatus_All"] <- last_entry$NestContentText
      }
    }
  }
  
  return(df)
}


phenology <- df_selected_BreedingSeason %>%
  find_first_E() %>%
  find_last_E() %>%
  find_first_C() %>%
  find_last_C() %>%
  find_last_visit()
```

## 6. Hatching Date Estimates 
The hatching dates were estimated based on the last egg visit entry and
the first chick visit entry, a mean of the two registered dates.
A new column for hatching date, hatching date accuracy, and hatching success is initialised. 
```{r}
phenology <- phenology %>%
  mutate(
    # Calculate Hatching_date as the mean of Last_E and First_C
    Hatching_date = if_else(
      !is.na(Last_E) & !is.na(First_C),
      as.Date((as.numeric(Last_E) + as.numeric(First_C)) / 2, origin = "1970-01-01"),
      NA_Date_
    ),
    
    # Calculate Hatching_date_accuracy as half the difference in days
    Hatching_date_accuracy = if_else(
      !is.na(Last_E) & !is.na(First_C),
      (as.numeric(First_C - Last_E) / 2),
      NA_real_
    ),
    
    # Set June 1st of the given BreedingSeason as Julian Date 1
    JulianHatchingDate = if_else(
      !is.na(Hatching_date),
      {
        # Extract the year from the Hatching_date
        year <- as.integer(format(Hatching_date, "%Y"))
        
        # Create June 1st date of the same year
        BreedingSeason_start <- as.Date(paste(year, "-06-01", sep = ""), format = "%Y-%m-%d")
        
        # Calculate Julian Date
        as.numeric(Hatching_date - BreedingSeason_start + 1)
      },
      NA_real_
    ),
    
    # Assign HatchingSuccess as 1 if Hatching_date is not NA, otherwise NA
    HatchingSuccess = if_else(
      !is.na(Hatching_date),
      1,
      NA_real_
    )
  )
```

## 7. Chick Presence and Breedingsuccess
```{r}

# Get the number of days a chick/chicks are present in the Nest 
NbChickPresence <- function(df) {
  # Ensure First_C, Last_C, Hatching_date, and LastOfVisitDate are Date objects
  df <- df %>%
    rowwise() %>%
    mutate(
      First_C = as.Date(First_C),
      Last_C = as.Date(Last_C),
      Hatching_date = as.Date(Hatching_date),
      LastOfVisitDate = as.Date(LastOfVisitDate)
    )
  
  # Group by NestNumber and BreedingSeason and calculate required metrics
  df <- df %>%
    group_by(NestNumber, BreedingSeason) %>%
    mutate(
      NbChickPresence = ifelse(!is.na(First_C) & !is.na(Last_C), as.numeric(Last_C - First_C + 1), NA),
      NbDays_ChickPresence_FromHD = ifelse(!is.na(Hatching_date) & !is.na(Last_C), as.numeric(Last_C - Hatching_date + 1), NA),
      Diff_LastVisitWithChick_LastVisit = ifelse(!is.na(Last_C) & !is.na(LastOfVisitDate), as.numeric(LastOfVisitDate - Last_C + 1), NA)
    ) %>%
    ungroup() # Ungroup to finalize the operation
  
  return(df)
}

# Apply the function
phenology <- NbChickPresence(phenology)


phenology <- phenology %>%
  rowwise() %>%
  mutate(
    Note = if_else(Hatching_date_accuracy > 5, "Hatching_date_accuracy > 5", Note),
    Note = if_else(NbDays_ChickPresence_FromHD < 15, "Entries for NbDays_ChickPresence_FromHD < 15", Note),
    ChickSurvival = if_else(Hatching_date_accuracy < 5 & any(NbDays_ChickPresence_FromHD > 15), 1, 0),
    BreedingSuccess = if_else(
      spcENG == "Glaucous gull" & ChickSurvival == 1, 1,  # Exception for Glaucous gull
      if_else(HatchingSuccess == 1 & ChickSurvival == 1, 1, 0)  # Regular condition
    )
  )

```

## 8. Final Phenology Dataset
```{r}
# Select the columns of interest
phenology <- phenology %>%
  select(spcENG, Locality, Area, NestNumber, BreedingSeason, Note, Last_E, First_C, Last_C, NbChickPresence, LastOfVisitDate, LastStatus_All, Hatching_date, JulianHatchingDate, Hatching_date_accuracy, NbDays_ChickPresence_FromHD, Diff_LastVisitWithChick_LastVisit, HatchingSuccess, ChickSurvival, BreedingSuccess)

view(phenology)

```
